---
title: "Stat5009_group17_rmd"
output: html_document
date: "2026-05-21"
---
Title: Detection of poor academics performance in students. 
It is complicated to detect student who are likely to fail, Some may be performing well all semester and at the end their performance declines. Vice Versa. Therefore, we are here to create a predictive model to detect students who may be in need of academic assistance early. 

The task: Supervised learning
Response (y): Final grade 
Features (x): 30 socio-economic features. 


```{r}
### Everyone needs to update this based on their paths and run accordingly 
setwd("D:/Curtin materials/stat5009_decision/A3/Stat5009_group17")
# install.packages("randomForest")
# install.packages("caret")
# for all random selection for repeatability
set.seed(75)
```

Merge data from both tables 
d1 is for maths results
d2 is for portugese results
d3 is for the data for intersection for students who are in both math and portugese
```{r}
# code given by lecturer to merge the datasets 
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(c("number of students: ", nrow(d3))) # 382 students
print("### Summary ###")
print(summary(d3))
```
- As we can see, the data has been cleaned, with no outliers and missing data. 

Print some data to see what they have
```{r}
head(d1)
head(d2)
head(d3)
``` 

convert the categorical data into numerical factors with hot encoding
```{r}
d1_encoded = data.frame(d1)
d1_encoded$school = ifelse(d1_encoded$school == "GP",0,1)
d1_encoded$sex = ifelse(d1_encoded$sex == "F",0,1)
d1_encoded$address = ifelse(d1_encoded$address == "R",0,1)
d1_encoded$famsize = ifelse(d1_encoded$famsize == "LE3",0,1)
d1_encoded$Pstatus = ifelse(d1_encoded$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d1_encoded$Mjob_teacher = ifelse(d1_encoded$Mjob == "teacher" , 1 , 0 )
d1_encoded$Mjob_health = ifelse(d1_encoded$Mjob == "health" , 1 , 0 )
d1_encoded$Mjob_services = ifelse(d1_encoded$Mjob == "services" , 1 , 0 )
d1_encoded$Mjob_athome = ifelse(d1_encoded$Mjob == "at_home" , 1 , 0 )
d1_encoded$Mjob_other = ifelse(d1_encoded$Mjob == "other" , 1 , 0 )
# remove the original column 
d1_encoded$Mjob = NULL

d1_encoded$Fjob_teacher = ifelse(d1_encoded$Fjob == "teacher" , 1 , 0 )
d1_encoded$Fjob_health = ifelse(d1_encoded$Fjob == "health" , 1 , 0 )
d1_encoded$Fjob_services = ifelse(d1_encoded$Fjob == "services" , 1 , 0 )
d1_encoded$Fjob_athome = ifelse(d1_encoded$Fjob == "at_home" , 1 , 0 )
d1_encoded$Fjob_other = ifelse(d1_encoded$Fjob == "other" , 1 , 0 )
# remove the original column 
d1_encoded$Fjob = NULL

d1_encoded$reason_home = ifelse(d1_encoded$reason == "home",1,0)
d1_encoded$reason_reputation = ifelse(d1_encoded$reason == "reputation",1,0)
d1_encoded$reason_course = ifelse(d1_encoded$reason == "course",1,0)
d1_encoded$reason_other = ifelse(d1_encoded$reason == "other",1,0)
# remove the original column 
d1_encoded$reason = NULL

d1_encoded$guardian_mother = ifelse(d1_encoded$guardian == "mother",1,0)
d1_encoded$guardian_father = ifelse(d1_encoded$guardian == "father",1,0)
d1_encoded$guardian_other = ifelse(d1_encoded$guardian == "other",1,0)
# remove the original column 
d1_encoded$guardian = NULL

d1_encoded$schoolsup = ifelse(d1_encoded$schoolsup == "yes", 1 ,0)

d1_encoded$famsup = ifelse(d1_encoded$famsup == "yes",1,0)

d1_encoded$paid =  ifelse(d1_encoded$paid == "no",0,1)
d1_encoded$activities =  ifelse(d1_encoded$activities == "no",0,1)
d1_encoded$nursery =  ifelse(d1_encoded$nursery == "no",0,1)
d1_encoded$higher =  ifelse(d1_encoded$higher == "no",0,1)
d1_encoded$internet =  ifelse(d1_encoded$internet == "no",0,1)
d1_encoded$romantic =  ifelse(d1_encoded$romantic == "no",0,1)

head(d1_encoded)
summary(d1_encoded)

# for d2
d2_encoded = data.frame(d2)
d2_encoded$school = ifelse(d2_encoded$school == "GP",0,1)
d2_encoded$sex = ifelse(d2_encoded$sex == "F",0,1)
d2_encoded$address = ifelse(d2_encoded$address == "R",0,1)
d2_encoded$famsize = ifelse(d2_encoded$famsize == "LE3",0,1)
d2_encoded$Pstatus = ifelse(d2_encoded$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d2_encoded$Mjob_teacher = ifelse(d2_encoded$Mjob == "teacher" , 1 , 0 )
d2_encoded$Mjob_health = ifelse(d2_encoded$Mjob == "health" , 1 , 0 )
d2_encoded$Mjob_services = ifelse(d2_encoded$Mjob == "services" , 1 , 0 )
d2_encoded$Mjob_athome = ifelse(d2_encoded$Mjob == "at_home" , 1 , 0 )
d2_encoded$Mjob_other = ifelse(d2_encoded$Mjob == "other" , 1 , 0 )
# remove the original column 
d2_encoded$Mjob = NULL

d2_encoded$Fjob_teacher = ifelse(d2_encoded$Fjob == "teacher" , 1 , 0 )
d2_encoded$Fjob_health = ifelse(d2_encoded$Fjob == "health" , 1 , 0 )
d2_encoded$Fjob_services = ifelse(d2_encoded$Fjob == "services" , 1 , 0 )
d2_encoded$Fjob_athome = ifelse(d2_encoded$Fjob == "at_home" , 1 , 0 )
d2_encoded$Fjob_other = ifelse(d2_encoded$Fjob == "other" , 1 , 0 )
# remove the original column 
d2_encoded$Fjob = NULL

d2_encoded$reason_home = ifelse(d2_encoded$reason == "home",1,0)
d2_encoded$reason_reputation = ifelse(d2_encoded$reason == "reputation",1,0)
d2_encoded$reason_course = ifelse(d2_encoded$reason == "course",1,0)
d2_encoded$reason_other = ifelse(d2_encoded$reason == "other",1,0)
# remove the original column 
d2_encoded$reason = NULL

d2_encoded$guardian_mother = ifelse(d2_encoded$guardian == "mother",1,0)
d2_encoded$guardian_father = ifelse(d2_encoded$guardian == "father",1,0)
d2_encoded$guardian_other = ifelse(d2_encoded$guardian == "other",1,0)
# remove the original column 
d2_encoded$guardian = NULL

d2_encoded$schoolsup = ifelse(d2_encoded$schoolsup == "yes", 1 ,0)

d2_encoded$famsup = ifelse(d2_encoded$famsup == "yes",1,0)

d2_encoded$paid =  ifelse(d2_encoded$paid == "no",0,1)
d2_encoded$activities =  ifelse(d2_encoded$activities == "no",0,1)
d2_encoded$nursery =  ifelse(d2_encoded$nursery == "no",0,1)
d2_encoded$higher =  ifelse(d2_encoded$higher == "no",0,1)
d2_encoded$internet =  ifelse(d2_encoded$internet == "no",0,1)
d2_encoded$romantic =  ifelse(d2_encoded$romantic == "no",0,1)

head(d2_encoded)
summary(d2_encoded)

# merge the code again
d3_encoded=merge(d1_encoded,d2_encoded,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob_teacher","Mjob_health","Mjob_services","Mjob_athome","Mjob_other","Fjob_teacher","Fjob_health","Fjob_services","Fjob_athome","Fjob_other","reason_home","reason_reputation","reason_course","reason_other","nursery","internet"))
 
```

linear regression for maths students
Does failing or absences related to the grades?
```{r}
# preparing the data for training and testing
n = nrow(d1_encoded)
training.d1_encoded.ind = sample(1:n, size = 0.8*n)

d1_encoded.training = d1_encoded[training.d1_encoded.ind,]
d1_encoded.test = d1_encoded[-training.d1_encoded.ind,]

# predicting if failing previously is correlated to final total
d1_encoded.fail.lm = lm(G3~failures, data=d1_encoded.training)
summary(d1_encoded.fail.lm)

print("#####################")
d1_encoded.absence.lm = lm(G3~absences, data=d1_encoded.training)
summary(d1_encoded.absence.lm)

print("#####################")
d1_encoded.study.lm = lm(G3~studytime, data=d1_encoded.training)
summary(d1_encoded.study.lm)
```

Trying auto multiple linear regression to see which fields are good
```{r}
library(leaps)
mod.fw = regsubsets(
  G3~.,
  nvmax = 8, 
  data = d1_encoded.training,
  method = "forward" 
  )
sm.mod.fw = summary(mod.fw)
print(sm.mod.fw)
```
As we can see G1 and G2 are highly related and used, which is not good because these are secondary response variables. So we have to recreate the forward model without these
```{r}
# remove the school and sex
d1_encoded.training.clean = d1_encoded.training[,-c(1,2)]
d1_encoded.test.clean = d1_encoded.test[, -c(1,2)]
mod.fw = regsubsets(
  G3~.
  -G1 -G2,
  nvmax = 8, 
  data = d1_encoded.training.clean,
  method = "exhaustive" 
  )
sm.mod.fw = summary(mod.fw)
print(sm.mod.fw)

# for bic, pick the lowest one
sm.mod.fw.bic = sm.mod.fw$bic

# for printing #
print(sm.mod.fw.bic)
ind = 1
for (x in 1:length(sm.mod.fw.bic) ){
  t = paste0("Model(",x ,"): ", sm.mod.fw.bic[x])
  print(t)
  
  if ( sm.mod.fw.bic[ind] > sm.mod.fw.bic[x]){
    ind = x
  }
  
}
t = paste0("lowest BIc model(",ind, "): ", sm.mod.fw.bic[ind])
print(t)


# higher is better for R^2
sm.mod.fw.r2 = sm.mod.fw$adjr2
print("R2")
print(sm.mod.fw.r2)
ind = 1
for (x in 1:length(sm.mod.fw.r2) ){
  t = paste0("Model(",x ,"): ", sm.mod.fw.r2[x])
  print(t)
  
  if ( sm.mod.fw.r2[ind] < sm.mod.fw.r2[x]){
    ind = x
  }
  
}
t = paste0("highest R^2 model(",ind, "): ", sm.mod.fw.r2[ind])
print(t)



```
So what are the features the model has selected?
- model 1: G3 ~ failures
- model 2: G3 ~ Medu + failures
- model 3: G3 ~ Mjob_services + Medu + failures
- model 4: G3 ~ romantic + Mjob_services + Medu + failures
- model 5: G3 ~ goout + romantic + Mjob_services + Medu + failures
- model 6: G3 ~ schoolsup + goout + romantic + Mjob_services + Medu + failures
- model 7: G3 ~ Mjob_teacher + Mjob_other + schoolsup + goout + romantic + Medu + failures
- model 8: G3 ~ reason_course + Mjob_teacher + Mjob_other + schoolsup + goout + romantic + Medu + failures
- model 9: G3 ~ Fjob_teacher +  reason_course + Mjob_teacher + Mjob_other + schoolsup + goout + romantic + Medu + failures
Hypothesis: 
1) Failures affect grades
2) Parent's being home and teaching children are a good indicator
What is the model with the lowest BIC? which is measure of goodness of fit against complexity, of -41.7615838856981
- model 3 which is  model 3: G3 ~ Mjob_services + Medu + failures

What is the model with the highest features explaining the variation in response of 0.209158023460924
- model 9, which is model 9: G3 ~ Fjob_teacher +  reason_course + Mjob_teacher + Mjob_other + schoolsup + goout + romantic + Medu + failures

So what is the better choice?
- model 3 has the lowest BIC of -41.7615838856981 with R^2 of 0.177531386039353,
while model 9 has BIC of -23.5607954630196, with R^2 of 0.203649356622243

-- so the transition from model 3 to 9 has an increase of R^2 of 0.02611797 while the BIC increased by 18.20079 

--- Therefore, the trade off between R^2 and BIC is "not worth it" and the model may have been learning noise instead. But let's assess further 


```{r}
# train the model on training data 
model3.lm = lm(G3 ~ Mjob_services + Medu + failures, data = d1_encoded.training.clean )
model9.lm = lm(G3 ~Fjob_teacher +  reason_course + Mjob_teacher + Mjob_other + schoolsup + goout + romantic + Medu + failures, data = d1_encoded.training.clean)

# evaluate on test data
preds_model3 = predict(model3.lm, newdata = d1_encoded.test)
preds_model9 = predict(model9.lm, newdata = d1_encoded.test)

actual_g3 = d1_encoded.test$G3

# calc mean square error
mse_m4 = mean((preds_model3-actual_g3)^2)
mse_m9 = mean((preds_model9-actual_g3)^2)

# calc residual mean square error
rmse_m4 = sqrt(mse_m4)
rmse_m9 = sqrt(mse_m9)

# print
output = data.frame(
  Model = c("model 3", "model 9"),
  r2 = c(0.177531386039353,0.203649356622243),
  mse = c(mse_m4,mse_m9),
  rmse = c(rmse_m4,rmse_m9)
)
print(output)

print(anova(model3.lm,model9.lm))
```
As we can see, model 3 has a lower MSE and RMSE. What does this mean? It means that on average, the squared difference between the prediction and actual is 17.35281  and the actual predictions deviate by approximately 4.165671 

With ANOVA, we can see that model9 has an F statistic of 2.7054  Which means that model9's ratio to of signal to noise is with a p value < 0.01 which means that this F statistics is very unlikely to be caused randomly. 

This means that with ANOVA, model9 is shown to be better. 

So there is a contradiction between our test MSE,RSME and ANOVA?
- this means that although the 5 new features added was good in training, it does not mean that it is good in generalization for future predictions. 
As it was deployed on test data, the MSE and RMSE increase which shows a degradation in performance. 

```{r}
print("## Model3")
print(summary(model3.lm))
```
Model4 interpretation
- G3 results will be 8.6672 on average if all features are 0 
- unit increase in Mjob_services, would lead to a increase in G3 results, 
-- because the P value is low, which means that the coefficient is statistically significant.
- unit increase in Medu would lead to a increase in G3 while being statistically significant 
-- this means that, if a mother is more educated, G3 is expected to increase
- unit increase in failure would lead to a decrease in G3 while being statistically significant 
-- therefore, it means that if a student fails more, it is more likely their grades will decrease. 
- the model only explains 17.75% of variations of G3, 
- the F statistics is the signal to noise ratio, is 23.66, means that the variance explained by the 3 features are 23 times bigger then the noise. Which is very strong
- p = 7.949e-14 which shows that the chance a signal this strong occurs by random is low. 

# creating Lasso and ridge
```{r}
library(glmnet)
x_train= model.matrix(G3~. -G1 -G2, data = d1_encoded.training.clean)[,-1]
y_train = d1_encoded.training.clean$G3

x_test= model.matrix(G3~. -G1 -G2, data = d1_encoded.test.clean)[,-1]
y_test = d1_encoded.test.clean$G3

# Lasso 
# k fold cross validation 
d1_cv_lasso.lm = cv.glmnet(x_train,y_train, alpha = 1)

# plot error path
plot(d1_cv_lasso.lm)

best_lambda_lasso = d1_cv_lasso.lm$lambda.min
sparse_lambda_lasso = d1_cv_lasso.lm$lambda.1se

d1_lasso_coef = predict(d1_cv_lasso.lm, type = "coefficient", s = best_lambda_lasso )
print(d1_lasso_coef)

# ridge
d1_cv_ridge.lm = cv.glmnet(x_train,y_train, alpha = 0)

# plot error path
plot(d1_cv_ridge.lm)

best_lambda_ridge = d1_cv_ridge.lm$lambda.min

d1_ridge_coef = predict(d1_cv_ridge.lm, type = "coefficient", s = best_lambda_ridge )
print(d1_ridge_coef)

# get stats
# lasso
d1_lasso_pred = predict(d1_cv_lasso.lm, s = best_lambda_lasso, newx = x_test)
lasso_mse = mean((d1_lasso_pred - y_test)^2)

d1_ridge_pred = predict(d1_cv_ridge.lm, s = best_lambda_ridge, newx = x_test)
ridge_mse = mean((d1_ridge_pred-y_test)^2)

# get the r^2 
# for glmnet the R^2 is DevRatio
# finding the index of where the best lambda came from 
lasso_index = which(d1_cv_lasso.lm$lambda == best_lambda_lasso)
ridge_index = which(d1_cv_ridge.lm$lambda == best_lambda_ridge)

# pull the best ratio
train_r2_lasso = d1_cv_lasso.lm$glmnet.fit$dev.ratio[lasso_index]
train_r2_ridge = d1_cv_ridge.lm$glmnet.fit$dev.ratio[ridge_index]
print(paste0("Lasso training r^2: ",train_r2_lasso))
print(paste0("Ridge training r^2: ",train_r2_ridge))
print(paste0("Lasso MSE: ", lasso_mse))
print(paste0("Ridge MSE: ", ridge_mse))
```
As we can see the ridge regression had the lower MSE and higher r^2. Likely due to the signals being weak and distributed such as Mjob. 

```{r}
# Convert the sparse matrix to a standard matrix layout
ridge_mat <- as.matrix(d1_ridge_coef)

print(ridge_mat)
```



Let's do some post check on the data
```{r}
# non linearity of actual data
# we are going to do residuals with model4 
# can also be done with residuals()
test_residuals = actual_g3 - preds_model3
plot(preds_model3,test_residuals)

test_residuals = actual_g3 - preds_model9
plot(preds_model9,test_residuals) 
```
No pattern is visible, hence true data can't be non linear, meaning the data is linear 


# Creating random forest
```{r}
library(randomForest)


rf = randomForest(G3~. -G1 -G2, data = d1_encoded.training.clean, importance = TRUE, ntree = 500 )
print(rf)

rf_pred = predict(rf, newdata = d1_encoded.test.clean)
rf_mse = mean((rf_pred - actual_g3)^2)
rf_rmse = sqrt(rf_mse)
print(paste0("MSE: ", rf_mse))
print(paste0("RMSE: ", rf_rmse))

# plot the error rate as the number of trees increase
plot(rf)
varImpPlot(rf)


```
Random forest had captured features to be 27.25% of the cause of variation in G3, with the lowest MSE and RMSE

```{r}
library(randomForest)
library(caret)

d1_encoded.training.clean.rf = data.frame(d1_encoded.training.clean)
d1_encoded.test.clean.rf = data.frame(d1_encoded.test.clean)

d1_encoded.training.clean.rf$final = as.factor(ifelse(d1_encoded.training.clean.rf$G3 >= 10, "Pass", "Fail"))
d1_encoded.test.clean.rf$final = as.factor(ifelse(d1_encoded.test.clean.rf$G3 >= 10, "Pass", "Fail"))

rf = randomForest(final~. -G1 -G2 -G3, data = d1_encoded.training.clean.rf, importance = TRUE, ntree = 500 )
print(rf)

rf_pred = predict(rf, newdata = d1_encoded.test.clean.rf)
rf_mse = mean((rf_pred - actual_g3)^2)
rf_rmse = sqrt(rf_mse)
print(paste0("MSE: ", rf_mse))
print(paste0("RMSE: ", rf_rmse))

# plot the error rate as the number of trees increase
plot(rf)
varImpPlot(rf)

# checking the classification accuracy
results = confusionMatrix(rf_pred, d1_encoded.test.clean.rf$final)
```


# final table
```{r}
model_name = c("Model 3 (subset linear regression, lowest BIC)",
               "Model 9 (subset linear regression, highest r^2)",
               "Lasso regression (regularized linear regression)",
               "Ridge regression (regularized linear regression)",
               "Random forest (non linear model)"
               )
train_r2 = c( 17.75, 20.36, 15.80, 21.70 , 27.84)

test_mse = c( 17.35,17.05, 16.09, 15.41, 15.35)

test_rmse = sqrt(test_mse)

scoreboard = data.frame(
  model = model_name,
  r2 = train_r2,
  mse = test_mse,
  rmse = test_rmse
)

print(scoreboard)

```

