---
title: "Stat5009_group17_rmd"
output: html_document
date: "2026-05-21"
---
Title: Detection of poor academics performance in students. 
It is complicated to detect student who are likely to fail, Some may be performing well all semester and at the end their performance declines. Vice Versa. Therefore, we are here to create a predictive model to detect students who may be in need of academic assistance early. 

The task: Supervised learning
Response (y): Final grade 
Features (x): 30 socio-economic features. 


```{r}
### Everyone needs to update this based on their paths and run accordingly 
setwd("D:/Curtin materials/stat5009_decision/A3/Stat5009_group17")
```

Merge data from both tables 
d1 is for maths results
d2 is for portugese results
d3 is for the data for intersection for students who are in both math and portugese
```{r}
# code given by lecturer to merge the datasets 
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(c("number of students: ", nrow(d3))) # 382 students
print("### Summary ###")
print(summary(d3))
```
- As we can see, the data has been cleaned, with no outliers and missing data. 

Print some data to see what they have
```{r}
head(d1)
head(d2)
head(d3)
``` 

convert the categorical data into numerical factors with hot encoding
```{r}
d1_encoded = data.frame(d1)
d1_encoded$school = ifelse(d1_encoded$school == "GP",0,1)
d1_encoded$sex = ifelse(d1_encoded$sex == "F",0,1)
d1_encoded$address = ifelse(d1_encoded$address == "R",0,1)
d1_encoded$famsize = ifelse(d1_encoded$famsize == "LE3",0,1)
d1_encoded$Pstatus = ifelse(d1_encoded$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d1_encoded$Mjob_teacher = ifelse(d1_encoded$Mjob == "teacher" , 1 , 0 )
d1_encoded$Mjob_health = ifelse(d1_encoded$Mjob == "health" , 1 , 0 )
d1_encoded$Mjob_services = ifelse(d1_encoded$Mjob == "services" , 1 , 0 )
d1_encoded$Mjob_athome = ifelse(d1_encoded$Mjob == "at_home" , 1 , 0 )
d1_encoded$Mjob_other = ifelse(d1_encoded$Mjob == "other" , 1 , 0 )
# remove the original column 
d1_encoded$Mjob = NULL

d1_encoded$Fjob_teacher = ifelse(d1_encoded$Fjob == "teacher" , 1 , 0 )
d1_encoded$Fjob_health = ifelse(d1_encoded$Fjob == "health" , 1 , 0 )
d1_encoded$Fjob_services = ifelse(d1_encoded$Fjob == "services" , 1 , 0 )
d1_encoded$Fjob_athome = ifelse(d1_encoded$Fjob == "at_home" , 1 , 0 )
d1_encoded$Fjob_other = ifelse(d1_encoded$Fjob == "other" , 1 , 0 )
# remove the original column 
d1_encoded$Fjob = NULL

d1_encoded$reason_home = ifelse(d1_encoded$reason == "home",1,0)
d1_encoded$reason_reputation = ifelse(d1_encoded$reason == "reputation",1,0)
d1_encoded$reason_course = ifelse(d1_encoded$reason == "course",1,0)
d1_encoded$reason_other = ifelse(d1_encoded$reason == "other",1,0)
# remove the original column 
d1_encoded$reason = NULL

d1_encoded$guardian_mother = ifelse(d1_encoded$guardian == "mother",1,0)
d1_encoded$guardian_father = ifelse(d1_encoded$guardian == "father",1,0)
d1_encoded$guardian_other = ifelse(d1_encoded$guardian == "other",1,0)
# remove the original column 
d1_encoded$guardian = NULL

d1_encoded$schoolsup = ifelse(d1_encoded$schoolsup == "yes", 1 ,0)

d1_encoded$famsup = ifelse(d1_encoded$famsup == "yes",1,0)

d1_encoded$paid =  ifelse(d1_encoded$paid == "no",0,1)
d1_encoded$activities =  ifelse(d1_encoded$activities == "no",0,1)
d1_encoded$nursery =  ifelse(d1_encoded$nursery == "no",0,1)
d1_encoded$higher =  ifelse(d1_encoded$higher == "no",0,1)
d1_encoded$internet =  ifelse(d1_encoded$internet == "no",0,1)
d1_encoded$romantic =  ifelse(d1_encoded$romantic == "no",0,1)

head(d1_encoded)
summary(d1_encoded)

# for d2
d2_encoded = data.frame(d2)
d2_encoded$school = ifelse(d2_encoded$school == "GP",0,1)
d2_encoded$sex = ifelse(d2_encoded$sex == "F",0,1)
d2_encoded$address = ifelse(d2_encoded$address == "R",0,1)
d2_encoded$famsize = ifelse(d2_encoded$famsize == "LE3",0,1)
d2_encoded$Pstatus = ifelse(d2_encoded$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d2_encoded$Mjob_teacher = ifelse(d2_encoded$Mjob == "teacher" , 1 , 0 )
d2_encoded$Mjob_health = ifelse(d2_encoded$Mjob == "health" , 1 , 0 )
d2_encoded$Mjob_services = ifelse(d2_encoded$Mjob == "services" , 1 , 0 )
d2_encoded$Mjob_athome = ifelse(d2_encoded$Mjob == "at_home" , 1 , 0 )
d2_encoded$Mjob_other = ifelse(d2_encoded$Mjob == "other" , 1 , 0 )
# remove the original column 
d2_encoded$Mjob = NULL

d2_encoded$Fjob_teacher = ifelse(d2_encoded$Fjob == "teacher" , 1 , 0 )
d2_encoded$Fjob_health = ifelse(d2_encoded$Fjob == "health" , 1 , 0 )
d2_encoded$Fjob_services = ifelse(d2_encoded$Fjob == "services" , 1 , 0 )
d2_encoded$Fjob_athome = ifelse(d2_encoded$Fjob == "at_home" , 1 , 0 )
d2_encoded$Fjob_other = ifelse(d2_encoded$Fjob == "other" , 1 , 0 )
# remove the original column 
d2_encoded$Fjob = NULL

d2_encoded$reason_home = ifelse(d2_encoded$reason == "home",1,0)
d2_encoded$reason_reputation = ifelse(d2_encoded$reason == "reputation",1,0)
d2_encoded$reason_course = ifelse(d2_encoded$reason == "course",1,0)
d2_encoded$reason_other = ifelse(d2_encoded$reason == "other",1,0)
# remove the original column 
d2_encoded$reason = NULL

d2_encoded$guardian_mother = ifelse(d2_encoded$guardian == "mother",1,0)
d2_encoded$guardian_father = ifelse(d2_encoded$guardian == "father",1,0)
d2_encoded$guardian_other = ifelse(d2_encoded$guardian == "other",1,0)
# remove the original column 
d2_encoded$guardian = NULL

d2_encoded$schoolsup = ifelse(d2_encoded$schoolsup == "yes", 1 ,0)

d2_encoded$famsup = ifelse(d2_encoded$famsup == "yes",1,0)

d2_encoded$paid =  ifelse(d2_encoded$paid == "no",0,1)
d2_encoded$activities =  ifelse(d2_encoded$activities == "no",0,1)
d2_encoded$nursery =  ifelse(d2_encoded$nursery == "no",0,1)
d2_encoded$higher =  ifelse(d2_encoded$higher == "no",0,1)
d2_encoded$internet =  ifelse(d2_encoded$internet == "no",0,1)
d2_encoded$romantic =  ifelse(d2_encoded$romantic == "no",0,1)

head(d2_encoded)
summary(d2_encoded)

# merge the code again
d3_encoded=merge(d1_encoded,d2_encoded,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob_teacher","Mjob_health","Mjob_services","Mjob_athome","Mjob_other","Fjob_teacher","Fjob_health","Fjob_services","Fjob_athome","Fjob_other","reason_home","reason_reputation","reason_course","reason_other","nursery","internet"))
 
```

linear regression for maths students
Does failing or absences related to the grades?
```{r}
# preparing the data for training and testing
n = nrow(d1_encoded)
training.d1_encoded.ind = sample(1:n, size = 0.8*n)

d1_encoded.training = d1_encoded[training.d1_encoded.ind,]
d1_encoded.test = d1_encoded[-training.d1_encoded.ind,]

# predicting if failing previously is correlated to final total
d1_encoded.fail.lm = lm(G3~failures, data=d1_encoded.training)
summary(d1_encoded.fail.lm)

print("#####################")
d1_encoded.absence.lm = lm(G3~absences, data=d1_encoded.training)
summary(d1_encoded.absence.lm)

print("#####################")
d1_encoded.study.lm = lm(G3~studytime, data=d1_encoded.training)
summary(d1_encoded.study.lm)
```

Trying auto multiple linear regression to see which fields are good
```{r}
library(leaps)
mod.fw = regsubsets(
  G3~.,
  nvmax = 8, 
  data = d1_encoded.training,
  method = "forward" 
  )
sm.mod.fw = summary(mod.fw)
print(sm.mod.fw)
```
As we can see G1 and G2 are highly related and used, which is not good because these are secondary response variables. So we have to recreate the forward model without these
```{r}
# remove the school and sex
d1_encoded.training.clean = d1_encoded.training[,-c(1,2)]
mod.fw = regsubsets(
  G3~.
  -G1 -G2,
  nvmax = 8, 
  data = d1_encoded.training.clean,
  method = "exhaustive" 
  )
sm.mod.fw = summary(mod.fw)
print(sm.mod.fw)

# for bic, pick the lowest one
sm.mod.fw.bic = sm.mod.fw$bic

# for printing #
print(sm.mod.fw.bic)
ind = 1
for (x in 1:length(sm.mod.fw.bic) ){
  t = paste0("Model(",x ,"): ", sm.mod.fw.bic[x])
  print(t)
  
  if ( sm.mod.fw.bic[ind] > sm.mod.fw.bic[x]){
    ind = x
  }
  
}
t = paste0("lowest BIc model(",ind, "): ", sm.mod.fw.bic[ind])
print(t)


# higher is better for R^2
sm.mod.fw.r2 = sm.mod.fw$adjr2
print("R2")
print(sm.mod.fw.r2)
ind = 1
for (x in 1:length(sm.mod.fw.r2) ){
  t = paste0("Model(",x ,"): ", sm.mod.fw.r2[x])
  print(t)
  
  if ( sm.mod.fw.r2[ind] < sm.mod.fw.r2[x]){
    ind = x
  }
  
}
t = paste0("highest R^2 model(",ind, "): ", sm.mod.fw.r2[ind])
print(t)



```
So what are the features the model has selected?
- model 1: G3 ~ failures
- model 2: G3 ~ Mjob_services + failures
- model 3: G3 ~ Mjob_athome + Mjob_other+ failures
- model 4: G3 ~ goout + Mjob_athome + Mjob_other+ failures
- model 5: G3 ~ romantic + goout + Medu + Mjob_services + failures
- model 6: G3 ~ freetime + romantic + goout + Medu + Mjob_services + failures
- model 7: G3 ~ famsup + freetime + romantic + goout + Medu + Mjob_services + failures
- model 8: G3 ~ Mjob_health + famsup + freetime + romantic + goout + Medu + Mjob_services + failures
- model 9: G3 ~ reason_reputation + reason_other + famsup + freetime + romantic + goout + Medu + Mjob_services + failures

What is the model with the lowest BIC? which is measure of goodness of fit against complexity, of -43.6781849770169
- model 4, which is  model 4: G3 ~ goout + Mjob_athome + Mjob_other+ failures

What is the model with the highest features explaining the variation in response of 0.237360925545721
- model 9, which is model 9: G3 ~ reason_reputation + reason_other + famsup + freetime + romantic + goout + Medu + Mjob_services + failures

So what is the better choice?
- model 4 has the lowest BIC of -43.6781849770169 with R^2 of 0.194678939518015,
while model 9 has the highest BIC of -37.2292841481151, with R^2 of 0.237360925545721

-- so the transition from model 4 to 9 has an increase of R^2 of 0.04268199 while the BIC increased by 6.448901 

--- Therefore, the trade off between R^2 and BIC is "not worth it" and the model may have been learning noise instead. But let's assess further 


```{r}
# train the model on training data 
model4.lm = lm(G3 ~ goout + Mjob_athome + Mjob_other+ failures, data = d1_encoded.training.clean )
model9.lm = lm(G3 ~ reason_reputation + reason_other + famsup + freetime + romantic + goout + Medu + Mjob_services + failures , data = d1_encoded.training.clean)

# evaluate on test data
preds_model4 = predict(model4.lm, newdata = d1_encoded.test)
preds_model9 = predict(model9.lm, newdata = d1_encoded.test)

actual_g3 = d1_encoded.test$G3

# calc mean square error
mse_m4 = mean((preds_model4-actual_g3)^2)
mse_m9 = mean((preds_model9-actual_g3)^2)

# calc residual mean square error
rmse_m4 = sqrt(mse_m4)
rmse_m9 = sqrt(mse_m9)

# print
output = data.frame(
  Model = c("model 4", "model 9"),
  r2 = c(0.194678939518015,0.237360925545721),
  mse = c(mse_m4,mse_m9),
  rmse = c(rmse_m4,rmse_m9)
)
print(output)

print(anova(model4.lm,model9.lm))
```
As we can see, model 4 has a lower MSE and RMSE. What does this mean? It means that on average, the squared difference between the prediction and actual is 21 and the actual predictions deviate by approximately 4.58. 

With ANOVA, we can see that model9 has an F statistic of 4.4811. Which means that model9's ratio to of signal to noise is 4.48 with a p value < 0.001 which means that this F statistics is very unlikely to be caused randomly. 

This means that with ANOVA, model9 is shown to be better. 

So there is a contradiction between our test MSE,RSME and ANOVA?
- this means that although the 5 new features added was good in training, it does not mean that it is good in generalization for future predictions.

