---
title: "Stat5009_group17_rmd"
output: html_document
date: "2026-05-21"
---
Title: Detection of poor academics performance in students. 
It is complicated to detect student who are likely to fail, Some may be performing well all semester and at the end their performance declines. Vice Versa. Therefore, we are here to create a predictive model to detect students who may be in need of academic assistance early. 

The task: Supervised learning
Response (y): Final grade 
Features (x): 30 socio-economic features. 


```{r}
### Everyone needs to update this based on their paths and run accordingly 
setwd("D:/Curtin materials/stat5009_decision/A3/Stat5009_group17")
```

Merge data from both tables 
d1 is for maths results
d2 is for portugese results
d3 is for the data for intersection for students who are in both math and portugese
```{r}
# code given by lecturer to merge the datasets 
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(c("number of students: ", nrow(d3))) # 382 students
print("### Summary ###")
print(summary(d3))
```
- As we can see, the data has been cleaned, with no outliers and missing data. 

Print some data to see what they have
```{r}
head(d1)
head(d2)
head(d3)
``` 

convert the categorical data into numerical factors with hot encoding
```{r}
d1$school = ifelse(d1$school == "GP",0,1)
d1$sex = ifelse(d1$sex == "F",0,1)
d1$address = ifelse(d1$address == "R",0,1)
d1$famsize = ifelse(d1$famsize == "LE3",0,1)
d1$Pstatus = ifelse(d1$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d1$Mjob = factor(d1$Mjob)
# to see the factors
levels(d1$Mjob)
d1$Fjob = factor(d1$Fjob)
d1$reason = factor(d1$reason)
d1$guardian = factor(d1$guardian)
d1$schoolsup = factor(d1$schoolsup)
d1$famsup = factor(d1$famsup)
d1$paid =  ifelse(d1$paid == "no",0,1)
d1$activities =  ifelse(d1$activities == "no",0,1)
d1$nursery =  ifelse(d1$nursery == "no",0,1)
d1$higher =  ifelse(d1$higher == "no",0,1)
d1$internet =  ifelse(d1$internet == "no",0,1)
d1$romantic =  ifelse(d1$romantic == "no",0,1)

head(d1)
summary(d1)

# for d2
d2$school = ifelse(d2$school == "GP",0,1)
d2$sex = ifelse(d2$sex == "F",0,1)
d2$address = ifelse(d2$address == "R",0,1)
d2$famsize = ifelse(d2$famsize == "LE3",0,1)
d2$Pstatus = ifelse(d2$Pstatus=="A",0,1)

# levels are to force it to follow our desired sequence
d2$Mjob = factor(d2$Mjob)
# to see the factors
levels(d2$Mjob)
d2$Fjob = factor(d2$Fjob)
d2$reason = factor(d2$reason)
d2$guardian = factor(d2$guardian)
d2$schoolsup = factor(d2$schoolsup)
d2$famsup = factor(d2$famsup)
d2$paid =  ifelse(d2$paid == "no",0,1)
d2$activities =  ifelse(d2$activities == "no",0,1)
d2$nursery =  ifelse(d2$nursery == "no",0,1)
d2$higher =  ifelse(d2$higher == "no",0,1)
d2$internet =  ifelse(d2$internet == "no",0,1)
d2$romantic =  ifelse(d2$romantic == "no",0,1)

# merge the code again
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))

```

linear regression for maths students
```{r}
# preparing the data for training and testing
n = nrow(d1)
training.d1.ind = sample(1:n, size = 0.8*n)

d1.training = d1[training.d1.ind,]
d1.test = d1[-training.d1.ind,]

# predicting if failing previously is correlated to final total
d1.fail.lm = lm(G3~failures, data=d1.training)
summary(d1.fail.lm)

print("#####################")
d1.absence.lm = lm(G3~absences, data=d1.training)
summary(d1.absence.lm)

print("#####################")
d1.study.lm = lm(G3~studytime, data=d1.training)
summary(d1.study.lm)
```

