---
title: "Stat5009_group17_rmd"
output: html_document
date: "2026-05-21"
---
Title: Detection of poor academics performance in students. 
It is complicated to detect student who are likely to fail, Some may be performing well all semester and at the end their performance declines. Vice Versa. Therefore, we are here to create a predictive model to detect students who may be in need of academic assistance early. 

The task: Supervised learning
Response (y): Final grade 
Features (x): 30 socio-economic features. 


```{r}
### Everyone needs to update this based on their paths and run accordingly 
setwd("D:/Curtin materials/stat5009_decision/A3/Stat5009_group17")
```

Merge data from both tables 
```{r}
# code given by lecturer to merge the datasets 
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(c("number of students: ", nrow(d3))) # 382 students
print("### Summary ###")
print(summary(d3))
```
- As we can see, the data has been cleaned, with no outliers and missing data. 

Print some data to see what they have
```{r}
head(d1)
head(d2)
``` 



